Poker Chips, Earthquakes, and Computation

### 2014-10-11 finaldraft.tex
finished computation section

\begin{figure}
\centering
\subfigure[Wire]{\includegraphics[width=0.25\linewidth]{wire.pdf}\label{fig:wire}}
\qquad\qquad
\subfigure[AND gate]{\includegraphics[width=0.25\linewidth]{andgate.pdf}\label{fig:AND}}
\qquad\qquad
\subfigure[OR gate]{\includegraphics[width=0.25\linewidth]{orgate.pdf}\label{fig:OR}}
\caption{Gates in the simple model. Green are the inputs, red are the outputs}
\label{fig:simplegates}
\end{figure}

@@

Our model of computation is as follows: we assume a finite board with finitely many chips. The inputs are designated by each two squares, one apart one is the ``clock" square and other the bit square. Likewise, the outputs each have a ``clock" square and a bit square. The ``clock" squares are set the value 4 and the bit squres are set to 4 if they are 1 and 0 otherwise. A computation is done by a series of earthquakes, when an output has the clock square active, then the bit square is read, and that output bit is considered read.

\subsection{Example}

A simple example is given by a board that computes the identity.

\[ \begin{array}{cccccc} 0 & 0 & 0 & 0 & 0 & 0 \\
				     4_c & 3 & 3 & 3 & 3 & 3_e \\
				     0 & 0 & 0 & 0 & 0 & 0 \\
				     i  & 3 & 3 & 3 & 3 & 3_o \\
				     0 & 0 & 0 & 0 & 0 & 0 \end{array} \]
Where the subscript $c$ indicates the clock, the subscript $e$ indicates the output clock square. $i$ indicates the bit square, and $o$ indicates the output bit.

One can see that $e$ will become active after 5 earthquakes since it will take that many earthquakes for $c$ to reach $e$. Also, if $i$ is initially inactive, then $o$ will be inactive, and if $i$ is active, then $o$ will be active. Therefore, we can compute the identity.

@@

Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance. this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $e$ is the output clock and $o$ is the output bit.

\[ \begin{array}{cccccc} 0 & 0 & e & 0 & o \\
				     0 & 3 & 2 & 0 & 3 \\
			            0 & 2 & 3 & 0 & 3 \\
				    0 & 3 & 2 & 0 & 3 \\
				    0 & 2 & 3 & 0 & 3 \\
				    c & 3 & 3 & 0 & 3 \\
				    0 & 0 & 0 & 0 & 3 \\
				    i & 3 & 3 & 3 & 3 \end{array} \]	
Since the clock snakes along the path while $i$ takes the shorter path. 
Likewise, we can perform a clockwise path by making $i$ snakes while the clock takes a straight path.

@@

We can synchronize the inputs so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
\[ \begin{array}{cccccccc} c & 3 & 3 & 3 & 3 & 3 & 3 & 0 \\
					 0 & 0 & 0 & 0 & 3 & 0 & 3 & 0 \\
					 i  & 3 & 3 & 3 & 1 & 3 & 3 & 3 \\
					 0 & 0 & 0 & 0 & 3 & 0 & 0 & 3 \\
					 0 & 0 & 0 & 0 & o & 0 & e & 3 \end{array} \]
Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long. $i$ waits for the clock since it needs the signal from $c$ from two sides to activate the 1 in the path from $i$ to $o$.

@@

So currently, we can compute any boolean circuit which can be drawn in a planar graph with only ANDs and ORs. So the question is whether we can compute the NOT? We argue that this is not possible.

\begin{theorem}
There does not exists any board configuration which computes the NOT operator in the above model of computation.
\end{theorem}

\begin{proof}
Suppose we have a board configuration $B$ with input squares $c$, $i$ and output squares $e, o$ that computes the NOT function. Then we can make another board $B'$ that synchronizes the final clock with the input so that we do not need to worry about timing of the clock. That is, we can guarantee that if $o$ is active, then $o$ is active at the same time $e$ is active.

Now lets make $B'$ an infinite board by putting 0s on the sides. So we have a finite board with finitely many chips. Let $S_{c_0i_0}$ be the set of squares which were active (had an earthquake affect them) during the execution of $B$ with the input value $i = i_0$ and $c = c_0$.

The first claim to make is that $S_{c_0i_0} \subset S_{c_1i_1}$ if $c_0 \leq c_1$ and $i_0 \leq i_1$. This is clearly true since the order of redistribution does not matter. In fact, we can do the earthquakes by first doing the earthquakes on $c$ and $i$ until the value decreases past $c_0$ and $i_0$, then simulate the computation of $c = c_0$ and $i = i_0$ with local earthquakes.

However, this arises in a contradiction, since $o \in S_{10}$ but $o \notin S_{11}$. Therefore, computing NOT with a board is not possible.
\end{proof}

What does this mean? Well, in particular, it means that we cannot do much. In particular, we can't add, since $x + 1$ will act as the NOT operator on the lowest order bit. We can only compute boolean circuits in planar graphs.

>>>>

\begin{figure}
\centering
\subfigure[Wire]{\includegraphics[width=0.25\linewidth]{wire.pdf}\label{fig:wire}}
\qquad\qquad
\subfigure[AND gate]{\includegraphics[width=0.25\linewidth]{andgate.pdf}\label{fig:AND}}
\qquad\qquad
\subfigure[OR gate]{\includegraphics[width=0.25\linewidth]{orgate.pdf}\label{fig:OR}}
\caption{Gates in the simple model. Green are the inputs, red are the outputs. We disregard edges and assume that any two nodes that are touching have an edge between them.}
\label{fig:simplegates}
\end{figure}

@@

As before, we assume a finite board with finitely many chips. The inputs are designated by having one input node. The outputs also each have one node. We now make the model of computation stronger by adding a special node $r$. When $r$ gets activated, it ends the execution and reads the outputs. We assume that in one ``round" we apply an earthquake to each valid node and that $r$ is the last node which gets applied an earthquake.

In order to make it easier to compute, we will make a special \textit{clock} node, designated as $c$ which always starts out with $4$ chips. We will have the clock go through each input and travel in parallel to each input. We will build \textit{gadgets} that allow us to compute with the clock so that we can build gates in a modular and then put them together.

We will also make a wire path that takes $c$ to $r$, so that we know exactly at what point the outputs are read.

As shown before, simple example is given by a board that computes the identity. For simplicity, we write the boards as a board of numbers, which every space is a node and two nodes have an edge between them if they are either above, below, left, or right.

\[ \begin{array}{cccccc} 0 & 0 & 0 & 0 & 0 & 0 \\
				     4_c & 3 & 3 & 3 & 3 & 3_r \\
				     0 & 0 & 0 & 0 & 0 & 0 \\
				     i  & 3 & 3 & 3 & 3 & 3_o \\
				     0 & 0 & 0 & 0 & 0 & 0 \end{array} \]
Where $i$ indicates the input bit node, and $o$ indicates the output bit.

One can see that $r$ will become active after 5 earthquakes since it will take that many earthquakes for $c$ to reach $r$. Also, if $i$ is initially inactive, then $o$ will be inactive, and if $i$ is active, then $o$ will be active. Therefore, we can compute the identity.

@@

Turning is non trivial, we must guarantee that we can turn a signal and align the timinings of the inputs and outputs in the clock so that they have the same distance. this is a counter-clockwise turn where $c$ denotes the clock, $i$ is the input bit, then $r$ is the output read and $o$ is the output bit.

\[ \begin{array}{cccccc} 0 & 0 & r & 0 & o \\
				     0 & 3 & 2 & 0 & 3 \\
			            0 & 2 & 3 & 0 & 3 \\
				    0 & 3 & 2 & 0 & 3 \\
				    0 & 2 & 3 & 0 & 3 \\
				    c & 3 & 3 & 0 & 3 \\
				    0 & 0 & 0 & 0 & 3 \\
				    i & 3 & 3 & 3 & 3 \end{array} \]	
Since the clock snakes along the path while $i$ takes the shorter path. 
Likewise, we can perform a clockwise path by making $i$ snakes while the clock takes a straight path.

@@

We can synchronize the input so that the input signal waits for the clock to arrive. This can be simply done with the following board configuration:
\[ \begin{array}{cccccccc} c & 3 & 3 & 3 & 3 & 3 & 3 & 0 \\
					 0 & 0 & 0 & 0 & 3 & 0 & 3 & 0 \\
					 i  & 3 & 3 & 3 & 1 & 3 & 3 & 3 \\
					 0 & 0 & 0 & 0 & 3 & 0 & 0 & 3 \\
					 0 & 0 & 0 & 0 & o & 0 & r & 3 \end{array} \]
Note that the path from $i$ to $o$ is 6 units long, while the path from clock is 12 unit long. $i$ waits for the clock since it needs the signal from $c$ from two sides to activate the 1 in the path from $i$ to $o$.

@@

We can build a NOT board with the following circuit:
\begin{figure}
\centering
\includegraphics[width=0.25\linewidth]{notgate.pdf}
\caption{Possible board to compute $f(x) = \neg x$}
\label{fig:NOT}
\end{figure}
In Figure~\ref{fig:NOT}, the wire to $c$ and $r$ is the same length as the wire from $c$ to $o$, but the wire from $i$ to $r$ is shorter than the length of the wire from $c$ to $o$. If $i$ is set to $1$, then $r$ will be activated before the clock reaches $o$, so $o$ will be read as 0. If $i$ is set to 0, then the clock will reach $o$ and $r$ at the same time, so the output is $1$. This computes $f(x) = \neg x$; however, we have switched the order of $r$ and $o$. We also do not know how to compose this board with other boards.

We would also like to build a crossover gadget to allow us to compute circuits which are not necesarily.
